White Dwarf Mergers on Adaptive Meshes\\ II. Nuclear Reactions

### 2017-09-19 papers/wdmerger_II/paper.tex
Simplify the timestep limiting discussion

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{e}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{X}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{e}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{X}$ 
determines the maximum change in the mass fraction of any species. To prevent species
whose abundances are small in absolute terms from controlling the timestep, we apply
the limiter based on $f_{X}$ only to species with abundances in a given zone greater
than $10^{-3}$ (this abundance threshold can be changed by the user of the code). By making 
$f_{e}$ and $f_{X}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_e$ parameter in the
literature is in the range of 0.2 or 0.3, while to our knowledge a limiter based on
$f_X$ has not been used by others performing these types of calculations. The sensitivity
of results to the value of these timestep limiters will be discussed in 
\autoref{sec:parameters:timestepping}. The factors $f_{e}$ and $f_{X}$ can be set at runtime in \castro.

At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum of all the
burning timesteps across all zones. To do this, we need a method for determining 
$\dot{e}$ and $\dot{X_k}$. A typical choice in the literature has been to set, for example,
\begin{equation}
  \dot{e} = \frac{e^{n} - e^{n-1}}{\Delta t^{n-1}}, \label{eq:burning_limiter_mode_4}
\end{equation}
where is $e^n$ is the internal energy at the start of the current timestep and
$e^{n-1}$ is the internal energy at the start of the previous timestep. 
The obvious analogue is used for constructing the species rate of change.
However, there are alternative methods of constructing this derivative estimate, 
and we have found that these different methods have measurable consequences.
We define four separate methods for calculating the time derivative, with 
the above being mode 4. Mode 3 is similar to mode 4 but replaces the
denominator in \autoref{eq:burning_limiter_mode_4} with the change in 
internal energy over the last timestep only from the nuclear reactions.
Mode 2 is the same as mode 3 but we only use the change in internal 
energy from the most recent nuclear burning step, that is, the second-half
of the Strang-split burning from the last timestep (the denominator 
then becomes $\Delta t / 2$). In mode 1, the most accurate option and 
the current default in \castro, we evaluate the right-hand-side of the 
burning network given the current state to explicitly obtain the 
instantaneous value of $\dot{e}$ and $\dot{X_k}$.

@@

To understand the consequences of this choice, and more broadly to 
understand the limitations of Strang splitting, we consider the 
basic outline of a single-level advance in an advection-reaction system:
\begin{enumerate}
  \item Evaluate timestep $\Delta t$ for the current advance
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Advance the hydrodynamics by $\Delta t$
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Return to Step 1
\end{enumerate}
Now, consider that during a head-on collision, initial nuclear burning 
will occur at the contact point between the two stars. Because of 
the staggered updates from splitting, the evolution effectively progresses 
as a cycle between burning for $\Delta t$ and getting fresh material 
advected into the contact point by the hydro update for $\Delta t$. 
When the collision begins, $\Delta t$ is controlled by the hydrodynamic 
stability criterion, and may be large enough that it is possible for 
the burning advance in Step 4 to completely burn the freshly advected 
material all the way to NSE. Consequently the evolution is no longer 
a good approximation to smooth burning of the in-falling material but
rather separate discrete burning and hydro steps, and the nature of 
the burning evolution will be quite different. Furthermore, by the 
time we return to Step 1 and estimate the next timestep size, all 
of the burning rates will be small again, and the instantaneous 
timestep limiter of mode 1 may actually substantially overestimate 
the needed timestep. The other modes will see that the energy/species  
substantially changed over the last timestep, but will still
overestimate the needed timestep because the burning was quiescent
for at least some portion of the last advance. Our experience has
shown that none of these methods is flawless, and that limiting
based on only changes in internal energy is particularly susceptible
to this staggered burning phenomenon; silicon-group material can
build up without changing the internal energy by a  large fraction,
so the timestep limiter is never triggered, and then in a single step
a substantial amount of iron-group elements can be be generated,
perhaps forming a detonation. In practice, however, we find that
this effect is not large enough to have a meaningful outcome, and
the results are largely insensitive to which method of limiting
is used.

With the timestep limited the way we advocate in this paper, 
the timesteps are generally short enough so that the errors 
due to splitting are small. Other approaches to the coupling 
between reactions and hydrodynamics have been proposed in the 
broader literature, especially iterative methods such as 
deferred corrections that allow each of these operators to 
feel the lagged effects of the other operators. For example,
in the context of low Mach number flows, \cite{nonaka:2012} have
used the method of spectral deferred corrections \citep{SDC} to
couple their advection-diffusion-reaction equation set. In our
context this would involve treating the full evolution equation
for each of the state variables as an ODE with a directly coupled
burning source term integrated at high order; the advective
flux is evaluated at the standard second order accuracy
and is included as an ODE source term.
We are presently investigating such a method,
and it may form the basis of further work on this subject.

Now we return to a point we hinted at above. The timestep
will only actually satisfy the energy criterion
$\Delta t \leq f_e \tau_e$ and species criterion
$\Delta t \leq f_{X_k} \tau_{X_k}$ when the estimates for
$\dot{e}$ and $\dot{X_k}$ we generate are at least as large
as the actual rate of change of energy and mass fractions
over the timestep. However, this can assumption can fail
during periods of runaway burning when the rate of change
of these quantities is highly nonlinear. We may not want
to neglect the errors caused by this approximation
because they may build up over an extended period of nonlinear
evolution and perhaps substantially change the final results.
To this end, we have implemented a timestep retry option in
\castro, which re-computes an advance if it violated the
stability criteria as judged from the end of the timestep.
We have found that the benefits for this problem are, however, small.

>>>>

In \castro, the reactions are coupled to the hydrodynamics using Strang splitting.
In a given timestep advance $\Delta t$, we first evolve the reactions alone through
a time interval $\Delta t / 2$. Then, we evolve the hydrodynamics for $\Delta t$,
and we evolve the reactions again for a further $\Delta t / 2$. The principal
drawback of this approach is that the reactions and the hydrodynamics can become
decoupled from each other. A common solution to this problem presented in
the literature has been to limit the size of the timestep and thereby limit the
extent of this decoupling \citep{raskin:2010,hawley:2012}, which we adopt here 
and have implemented in \castro. Defining the nuclear energy injection timescale 
$\tau_e$, and the species evolution timescale $\tau_{X_k}$,
\begin{align}
  \tau_e &\equiv \frac{e}{|\dot{e}|} \\
  \tau_{X_k} &\equiv \frac{X_k}{|\dot{X_k}|},
\end{align}
where $\dot{e}$ is an estimate of the time rate of change of the internal energy
from nuclear burning, and $\dot{X_k}$ is an estimate of the time rate of change 
of the mass fraction of the species with index $k$, we define burning-limited 
timesteps $\Delta t_{be}$ and $\Delta t_{bX_k}$:
\begin{align}
  \Delta t_{be} &= f_{be}\, \tau_e \label{eq:timestep_e}\\
  \Delta t_{bX_k} &= f_{bX}\, \tau_{X_k}. \label{eq:timestep_X}
\end{align}
Given an estimate for $\dot{e}$, the factor $f_{be}$ determines by what 
fraction we would like to allow the internal energy to change
in the current timestep, under the assumption that $\dot{e}$ does not change from
timestep to timestep. Similarly, given an estimate for $\dot{X_k}$, the factor $f_{bX}$ 
determines the maximum change in the mass fraction of any species. To prevent species
whose abundances are small in absolute terms from controlling the timestep, we apply
the limiter based on $f_{bX}$ only to species with abundances in a given zone greater
than $10^{-3}$ (this abundance threshold can be changed by the user of the code). By making 
$f_{be}$ and $f_{bX}$ smaller, we can control the magnitude of the decoupling 
between the reactions and the hydro. A typical choice for $f_{be}$ in the
literature is in the range of 0.1 to 0.3, while to our knowledge a limiter based on
$f_{bX}$ has not been used by others performing these types of calculations. The sensitivity
of results to the value of these timestep limiters will be discussed in 
\autoref{sec:temporalresolution}. The factors $f_{be}$ and $f_{bX}$ can be set at runtime in \castro.

At the start of each advance, we limit the size of the timestep to be the smaller
of the minimum hydro timestep (limited by the CFL condition), and the minimum of all the
burning timesteps across all zones. To do this, we need a method for determining 
$\dot{e}$ and $\dot{X_k}$. We choose to do this with a direct evaluation of the
right-hand-side of the burning network, given the current state at the time when
the timestep is evaluated. Other work in the literature has typically evaluated
this as a finite difference using the change over the last timestep. Our method
is more accurate in principle, but in practice for this problem we find no
meaningful difference between the limiting we propose and other methods of
estimating the time derivatives.

To be precise, we note that the timestep
will only actually satisfy the energy criterion
$\Delta t \leq f_e \tau_e$ and species criterion
$\Delta t \leq f_{X_k} \tau_{X_k}$ when the estimates for
$\dot{e}$ and $\dot{X_k}$ we generate are at least as large
as the actual rate of change of energy and mass fractions
over the timestep. However, this can assumption will sometimes fail
during periods of runaway burning when the rate of change
of these quantities is highly nonlinear. We may not want
to neglect the errors caused by this approximation
because they may build up over an extended period of nonlinear
evolution and perhaps substantially change the final results.
To this end, we have implemented a timestep retry option in
\castro, which re-computes an advance if it violated the
stability criteria as judged from the end of the timestep,
using subcycled timesteps which individually satisfy the
stability criterion to the tolerance we have asked for.
We have found that the benefits for this problem are, however, small.

@@

To understand the consequences of the timestep limiting, and more broadly to 
understand the limitations of Strang splitting, we consider the 
basic outline of a single-level advance in an advection-reaction system:
\begin{enumerate}
  \item Evaluate timestep $\Delta t$ for the current advance
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Advance the hydrodynamics by $\Delta t$
  \item Advance the nuclear burning network by $\Delta t / 2$
  \item Return to Step 1
\end{enumerate}
Now, consider that during a head-on collision, initial nuclear burning 
will occur at the contact point between the two stars. Because of 
the staggered updates from splitting, the evolution effectively progresses 
as a cycle between burning for $\Delta t$ and getting fresh material 
advected into the contact point by the hydro update for $\Delta t$. 
When the collision begins, $\Delta t$ is controlled by the hydrodynamic 
stability criterion, and may be large enough that it is possible for 
the burning advance in Step 4 to completely burn the freshly advected 
material all the way to NSE. Consequently the evolution is no longer 
a good approximation to smooth burning of the in-falling material but
rather separate discrete burning and hydro steps, and the nature of 
the burning evolution will be quite different. As the timestep becomes
smaller, the evolution becomes smoother and the splitting error decreases.

Other approaches to the coupling 
between reactions and hydrodynamics have been proposed in the 
broader literature, especially iterative methods such as 
deferred corrections that allow each of these operators to 
feel the lagged effects of the other operators. For example,
in the context of low Mach number flows, \cite{nonaka:2012} have
used the method of spectral deferred corrections \citep{SDC} to
couple their advection-diffusion-reaction equation set. In our
context this would involve treating the full evolution equation
for each of the state variables as an ODE with a directly coupled
burning source term integrated at high order; the advective
flux is evaluated at the standard second order accuracy
and is included as an ODE source term.
We are presently investigating such a method,
and it may form the basis of further work on this subject.
